---
title: "Tube usage"
author: "Rob Wells"
date: "2018-04-17 21:00"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

We’ll need Hadley Wickham’s libraries, so we’ll use the shortcut to bring the the core Tidyverse in, as well as a date-handling library. We’ll also set the default plot theme.

```{r message=FALSE, warning=FALSE}
library(tidyverse)
library(lubridate)

theme_rjw <- function(base_size = 14, base_family = 'Source Sans Pro')
{
    theme_minimal(base_size = base_size,
                  base_family = base_family) %+replace%
        theme(panel.background = element_blank(),
              plot.background = element_rect(
                  fill = '#f7f8f9', color = NA),
              legend.background = element_blank()) +
        theme(plot.title = element_text(margin = margin(b = 5)),
              plot.subtitle = element_text(margin = margin(b = 5)),
              axis.title.x.top = element_text(margin = margin(t = 5, b = 10)),
              axis.text = element_text(color = '#607080'),
              panel.grid.major = element_line(color = '#60708040'),
              panel.grid.minor = element_line(color = '#60708033'))
}

theme_set(theme_rjw())

mav <- function(x,n=5){stats::filter(x,rep(1/n,n), sides = 1)}
```


All of the journey history data is stored in the original CSV files emailed to me by Transport for London. At this point I’ve accrued quite a few.

```{r}
oyster_filenames <- dir(
    '~/Documents/Oyster card/Journey history CSVs/',
    pattern = '*.csv', full.names = TRUE)

length(oyster_filenames)
```

Now we need to read in the data, but importantly we need to skip the initial header row and then combine all of the per-CSV data into one frame.

```{r message=FALSE}
oyster_data <- oyster_filenames %>%
    map(~ read_csv(., skip = 1)) %>%
    reduce(rbind)
```

The funky `%>%` operator pipes the data through, just like at the Unix command line. Using `map` and `~` allow us to specify arguments to `read_csv`, to skip the column headers. The last part of this pipeline reduces the 100-odd frames into a single one.

Let’s take a look at our data before we start to clean it up.

```{r}
nrow(oyster_data)  # Number of rows
head(oyster_data)
```

Just from the top few rows, you can spot that the data is untidy. Line 3 shows a top-up to the balance on the card being picked up, while the rest are records of journeys made.

The date and time columns need combining. It’s unclear what we need out of Charge, Credit and Balance at the moment, but clearly not all three. There’s also a note column, but it’s mostly empty. (NA is R’s missing value.)

```{r}
sum( !(oyster_data$Note %>% is.na) ) # Number of non-NA Note fields
oyster_data %>%
    select(Note) %>%
    distinct
```

This is a warning of some data-quality problems: no start or ending station for some journeys. We can safely leave rows mentioning fare capping, as well as journey continuation ([which is fairly new][hopper]).

[hopper]: https://www.londonreconnections.com/2016/dont-fear-beeper-bus-hopper-tickets-future-oyster/

While we mull over how or whether to handle the incomplete journey records, let’s see what might need doing to incomplete journeys where I’ve been able to say where I tapped in or out afterwards.

```{r warning=FALSE}
oyster_data %>%
    filter( Note %>% startsWith('This incomplete') ) %>%
    select('Journey/Action')
```

Angle brackets around the station that was added after-the-fact. Let’s see if any other rows contain angle brackets. (`grepl` returns a logical vector that `filter` can make use of.)

```{r}
oyster_data %>%
    filter( grepl('[<>]', `Journey/Action`) ) %>%
    nrow
```

Nope, OK. Let’s just trim out those brackets. (There must be a tidier way of doing this.)

```{r}
bracket_row <- grep('[<>]', oyster_data$`Journey/Action`)
bracket_text <-  oyster_data$`Journey/Action`[bracket_row]
oyster_data$`Journey/Action`[bracket_row] <- 
    gsub('[<>]', '', bracket_text)
print(oyster_data$`Journey/Action`[bracket_row])
```

Next, let’s find out what else is lurking in our data. We’ve already seen rail journeys, which start and end (ideally) at a station. But is there anything else in this data frame that doesn’t match that formulation?

```{r}
oyster_data %>%
    select('Journey/Action') %>%
    filter( !grepl('.+ to .+', `Journey/Action`) )
```

There’s still quite a few of these, but there’s a pattern: the activity type, then a comma, then more detail. What kinds of activity are we dealing with?

```{r}
unique((oyster_data %>%
    select('Journey/Action') %>%
    filter( !grepl('.+ to .+', `Journey/Action`) ) %>%
    map(~ gsub('^([^,]+),.+$', '\\1', .))
)$`Journey/Action`)
```

Things got a little weird there, with me somehow creating a list with a sole property of the name I selected from the frame. But hey ho, I’m still learning the ropes with R.

Anyway, as far as travel goes, the bus journeys are the only interesting type here. (Well, the “Unknown transaction” rows caused me a little bit of concern! But it seems to tally, though still oddly, with my old commute.)

```{r}
oyster_data %>%
    filter( grepl('Unknown transaction', `Journey/Action`) ) %>%
    select(Date, `Start Time`, `Journey/Action`, `Charge`, `Credit`)
```

Anyway, we can draw up a list of row prefixes that we’ll exclude from the data we use for analysis:

* Auto top-up
* Automated Refund
* Oyster helpline refund
* Unknown transaction
* Entered and exited Woolwich Arsenal DLR

Let’s also jettison journeys where we don’t know the start or end station. Thankfully there are only a few of these.

```{r}
oyster_data %>%
    filter(Note %>% startsWith('We are not able to show')) %>%
    nrow
```

I’m also going to exclude bus journeys for the same reason, to make things simpler. I rarely take the bus so it’s no great loss.

```{r}
oyster_data %>%
    filter(`Journey/Action` %>% startsWith('Bus journey')) %>%
    nrow
```

Now, all but one of the cases mentioned above can be excluded by filtering for missing start or end times. The oddity case in when I entered and then immediately left Woolwich Arsenal, so we’ll handle that specifically.

```{r}
working_data <- oyster_data %>%
    filter( !grepl('Entered and exited', `Journey/Action`)) %>%
    drop_na(`Start Time`, `End Time`)
```

Let’s look at sorting out our dates. Really the start and finish times should be a datetime type, rather than times stored alongside a date. But there’s a possibility that the finish time journeys will actually be the following calendar day (after midnight).

```{r}
working_data <- working_data %>%
    mutate(
        start_dt = dmy_hms(paste(Date, `Start Time`),tz = 'Europe/London'),
        ends_next_day = `End Time` < `Start Time`,
        end_dt = dmy_hms(paste(Date, `End Time`), tz = 'Europe/London') + days(1 * ends_next_day))
working_data %>%
    select(start_dt, end_dt) %>%
    head
```

Great. Now let’s split up the stations into where I entered and exited the network.

```{r}
working_data <- working_data %>%
    mutate(
        enter = gsub('(.+) to (.+)', '\\1', `Journey/Action`),
        exit = gsub('(.+) to (.+)', '\\2', `Journey/Action`),
    )
working_data %>%
    select(enter, exit) %>%
    head
```

Fab. Let’s just have a quick look at the remaining journeys that have some kind of note attached to them, out of interest, excluding the corrected-incomplete journey from earlier.

```{r}
remaining_notes <- working_data %>%
    filter(!is.na(Note), !grepl('incomplete', Note)) %>%
    arrange(start_dt) %>%
    select(start_dt, enter, exit, Charge, Note)
```

They’re all journeys where the fare reduced because I’d hit the cap for the day, but fewer than 20 across 3.5 years.

Let’s at long last put our data into a tidy format.

```{r}
tidied <- working_data %>%
    select(start = start_dt,
           finish = end_dt,
           enter, exit,
           fare = Charge) %>%
    arrange(start)
head(tidied)
```

After much faffing about, we have the basics we want, and correctly converted. The journey start and finish times and stations, and the fare.


### Plots

Let’s now look at getting this reasonably clean data into some pretty pictures. To warm up, we’ll just plot the cumulative spending over the period.

```{r}
ggplot(data = tidied) +
    geom_vline(xintercept = as.POSIXct('2016-08-01'),
               alpha = 0.5) +
    annotate(geom = 'text',
             x = as.POSIXct('2017-01-15'), y = 325,
             label = '← Moved nearer to work',
             family = 'Source Sans Pro') +
    geom_line(mapping = aes(x = start, y = cumsum(fare))) +
    scale_x_datetime(date_breaks = '6 months',
                     date_labels = '%b %Y',
                     limits = as.POSIXct(c(ymd('2014-09-06'),
                                           ymd('2018-04-25'))),
                     expand = c(0, 0)) +
    scale_y_continuous(labels = scales::dollar_format(prefix = '£'),
                       breaks = c(0, 500, 1000, 1400, 1650)) +
    labs(x = NULL, y = NULL,
         title = 'Cumulative spending on London rail travel',
         subtitle = 'September 6 2014 to April 25 2018')
```

Now let’s see if we can recreate the first plot I showed the first time I ran this kind of analysis, weekly totals with a running average.

```{r}
weekly_fares <- tidied %>%
    group_by(year_week = format(start, '%G %V')) %>%
    summarise(week_end = max(start),
              week_total = sum(fare))

ggplot(weekly_fares, mapping = aes(x = week_end, y = week_total)) +
    geom_point() +
    geom_smooth(method = 'loess') +
    theme_rjw() +
    scale_x_datetime(date_breaks = '6 months',
                     date_labels = '%b %Y',
                     limits = as.POSIXct(c(ymd('2014-09-01'),
                                           ymd('2018-06-01'))),
                     expand = c(0, 0)) +
    scale_y_continuous(labels = scales::dollar_format(prefix = '£'),
                       breaks = c(0, 10, 20, 30, 40),
                       limits = c(0, 40),
                       expand = c(0, 0))
```

But this isn’t ideal, because we’re missing weeks where I didn’t travel at all. So we need to add in those zero-expenditure weeks. Let’s create a sequence of weeks starting from the first in the data frame above, and then join that against the real data, keeping the zero values when there isn’t any data.

```{r warning=FALSE}
week_series <- seq(ymd('2014-09-07'),
                   ymd('2018-04-25'),
                   by = '1 week') %>%
    tibble(
        week_end = as_datetime(.),
        year_week = format(., '%G %V'),
        week_total = 0)

week_series <- week_series %>%
    select(year_week, week_end, week_total)

joined_weeks <- full_join(week_series, weekly_fares,
                          by = 'year_week') %>%
    mutate(week_end = coalesce(week_end.y, week_end.x),
           week_total = coalesce(week_total.y, week_total.x)) %>%
    select(-contains('.'))

ggplot(joined_weeks, mapping = aes(x = week_end, y = week_total)) +
    geom_vline(xintercept = as.POSIXct('2016-08-01'),
               alpha = 0.5) +
    geom_point(size = .5, color = '#1369bf') +
    geom_line(mapping = aes(x = week_end, y = mav(week_total, 6)),
              color = '#bd3338', size = .75) +
    scale_x_datetime(date_breaks = '6 months',
                     date_labels = '%b ’%y',
                     limits = as.POSIXct(c(ymd('2014-09-01'),
                                           ymd('2018-06-01'))),
                     expand = c(0, 0)) +
    scale_y_continuous(labels = scales::dollar_format(prefix = '£'),
                       breaks = c(0, 10, 20, 30, 40),
                       limits = c(0, 40)) +
    labs(x = NULL,
         y = 'Weekly total cost',
         title = 'Weekly transport spending and six-week moving average',
         subtitle = 'September 6 2014 to April 25 2018; vertical line marks house move',
         caption = 'This isn’t tremendously informative')
```

The number of weeks where I don’t travel at all increases notably after I moved house. Let’s focus in on that for a second.

```{r fig.width=8, fig.height=2}
joined_weeks %>%
    ggplot(mapping = aes(x = year_week, y = 1,
                         alpha = week_total == 0)) +
    geom_col(width = 1) +
    scale_y_discrete(limits = c(FALSE, TRUE),
                     labels = NULL,
                     breaks = NULL,
                     expand = c(0, 0)) +
    scale_x_discrete(breaks = NULL) +
    scale_alpha_discrete(labels = c('Some', 'Zero')) +
    labs(title = 'Weeks with no journeys made',
         subtitle = 'September 6 2014 to April 25 2018',
         x = NULL, y = NULL,
         alpha = 'Number of journeys') +
    theme(legend.position = 'bottom')
```

Now let’s see if we can plot a moving average of the data:

```{r warning=FALSE}
ggplot(joined_weeks) +
    geom_vline(xintercept = as.POSIXct('2016-08-01'),
               alpha = 0.5) +
    geom_line(mapping = aes(x = week_end, y = mav(week_total, 13))) +
    scale_x_datetime(date_breaks = '6 months',
                     date_labels = '%b %Y',
                     limits = as.POSIXct(c(ymd('2014-09-01'),
                                           ymd('2018-06-01'))),
                     expand = c(0, 0)) +
    scale_y_continuous(labels = scales::dollar_format(prefix = '£'),
                       breaks = c(0, 10, 20, 30, 40),
                       limits = c(0, 40),
                       expand = c(0, 0))
```

Let’s try to plot a moving average of zero-spending weeks:

```{r warning=FALSE}
ggplot(joined_weeks) +
    geom_vline(xintercept = as.POSIXct('2016-08-01'),
               alpha = 0.5) +
    geom_line(aes(x = week_end, y = mav(week_total == 0, 26))) +
    scale_x_datetime(date_breaks = '6 months',
                     date_labels = '%b %Y',
                     limits = as.POSIXct(c(ymd('2014-09-01'),
                                           ymd('2018-06-01'))),
                     expand = c(0, 0)) +
    scale_y_continuous(limits = c(0, 1),
                       labels = scales::percent_format(),
                       expand = c(0, 0))
```

Now let’s look at the length of journeys:

```{r}
ggplot(tidied) +
    geom_histogram(aes(finish - start),
                   breaks = seq(0, 80, 5),
                   fill = 'royalblue',
                   closed = 'left',  # Important for matching previous
                   colour = 'black',
                   size = 0.2) +
    scale_x_continuous(limits = c(0, 80),
                       expand = c(0, 0),
                       breaks = seq(0, 80, 10),
                       minor_breaks = NULL,
                       labels = seq(0, 80, 10)) +
    scale_y_continuous(breaks = seq(0, 350, 50),
                       minor_breaks = NULL,
                       labels = seq(0, 350, 50),
                       limits = c(0, 350),
                       expand = c(0, 0))
```

So far, that looks like that histogram in my original post about this topic. Let’s plot the difference between journey lengths when I lived in Woolwich and after my move to Leyton:

```{r warning=FALSE}
ggplot(tidied) +
    geom_histogram(aes(x = finish - start,
                       fill = start > ymd('20160801')),
                   breaks = seq(0, 80, 5),
                   closed = 'left',  # Important for matching previous
                   colour = 'black',
                   size = 0.2) +
    scale_x_continuous(limits = c(0, 80),
                       expand = c(0, 0),
                       breaks = seq(0, 80, 10),
                       minor_breaks = NULL,
                       labels = seq(0, 80, 10)) +
    scale_y_continuous(breaks = seq(0, 350, 50),
                       minor_breaks = NULL,
                       labels = seq(0, 350, 50),
                       limits = c(0, 350),
                       expand = c(0, 0)) +
    scale_fill_discrete(labels = c('Before move', 'After move'),
                        position = 'top') +
    labs(title = 'Journey length',
         subtitle = 'Five-minute groups, bars stacked',
         x = 'Duration (minutes)',
         y = 'Number of journeys',
         fill = NULL) +
    theme(legend.position = 'bottom')
```

You can see that my post-move journeys tend to be shorter; this is likely because I can now easily get the high-speed train from Stratford International to St Pancras (for national rail services to, and from, my home town of Leicester). The high-speed journey itself is about 7 minutes, so a bit of waiting around at each end could easily push it up into the 10-minute bin.

```{r warning=FALSE}
paired <- tidied %>%
    rowwise %>%
    mutate(station_pair = paste(sort(c(enter, exit)),
                                collapse = '\n'),
           after_move = start > ymd('20160801'))
top_pairs <- paired %>%
    group_by(station_pair) %>%
    summarise(count = n()) %>%
    filter(count >= 10)

# Bring the counts into the frame to make life easier
subsetted <- subset(paired, station_pair %in% top_pairs$station_pair) %>%
    left_join(top_pairs, by = 'station_pair')

station_cleaner <- function(x) {
    x %>%
        str_replace_all(' DLR', '') %>%
        str_replace_all('International', 'Int’l') %>%
        str_replace_all(' \\[London [^\n]+', '')
}

ggplot(subsetted) +
    geom_bar(aes(x = reorder(station_pair, count),
                 fill = after_move),
             stat = 'count',
             position = 'stack',
             width = 0.8) +
    coord_flip() +
    scale_x_discrete(labels = station_cleaner) +
    scale_y_continuous(breaks = seq(0, 300, 50),
                       minor_breaks = NULL) +
    scale_fill_discrete(labels = c('Before move', 'After move')) +
    labs(title = 'Most common journeys',
         subtitle = 'In either direction between stations, 10 or more trips (bars stacked)',
         y = NULL, x = NULL, fill = NULL) +
    theme(legend.position = 'bottom',
          panel.grid.major.y = element_blank(),
          axis.text.y = element_text(
              size = 9,
              margin = unit(c(0, -4, 0, 0), 'mm')))
```

```{r warning=FALSE}
ggplot(tidied, aes(x = start,
                   y = mav(fare, 10))) +
    geom_vline(xintercept = as.POSIXct('2016-08-01'),
               alpha = 0.5) +
    annotate(geom = 'rect',
             xmin = as.POSIXct('2015-03-20'),
             xmax = as.POSIXct('2016-02-05'),
             ymin = 0.95, ymax = 1.40,
             alpha = 0.2, fill = '#1369bf') +
    annotate(geom = 'text',
             x = as.POSIXct('2016-11-14'),
             y = 1.10,
             label = '← Off-peak railcard discount applied',
             family = 'Source Sans Pro') +
    geom_point(size = .75) +
    scale_y_continuous(labels = scales::dollar_format(prefix = '£')) +
    labs(x = NULL,
         y = 'Fare',
         title = 'Journey cost over time',
         subtitle = '10-trip moving average. Vertical line marks house move')
```

```{r warning=FALSE}
ggplot(tidied, aes(x = start,
                   y = mav(finish - start, 10))) +
    geom_vline(xintercept = as.POSIXct('2016-08-01'),
               alpha = 0.5) +
    geom_point(size = 0.75) +
    scale_y_continuous() +
    labs(x = NULL,
         y = 'Duration (minutes)',
         title = 'Journey length over time',
         subtitle = '10-trip moving average. Vertical line marks house move')
```


```{r warning=FALSE}
ggplot(tidied, aes(x = fare,
                   y = finish - start,
                   colour = start > ymd('20160801'))) +
    geom_point(alpha = 0.5) +
    facet_wrap(~ start > ymd('20160801'),
               labeller = as_labeller(
                   c(`TRUE` = 'After move', `FALSE` = 'Before move'))) +
    scale_y_continuous(limits = c(0, 80)) +
    scale_x_continuous(labels = scales::dollar_format(prefix = '£'),
                       limits = c(0, 6)) +
    labs(x = 'Fare',
         y = 'Journey duration (minutes)',
         title = 'Journey cost and duration',
         subtitle = 'Are post-move journeys shorter and more expensive?') +
    guides(colour = FALSE) +
    theme(panel.border = element_rect(colour='black', fill=NA))
```

Now let’s go back to time and see when my journeys were most commonly made, and the difference before and after the move.

```{r}
tidied %>%
    mutate(start_hour = case_when(
        hour(start) == 0 ~ as.integer(24),
        TRUE ~ hour(start)
    )) %>%
    # filter(start > ymd('20160801')) %>%
    ggplot(aes(x = start_hour,
               y = ..density..,
               fill = start > ymd('20160801'))) +
    geom_histogram(binwidth = 1,
                   closed = 'left',
                   color = 'black',
                   size = 0.2) +
    facet_wrap(~ start > ymd('20160801'),
               ncol = 1,
               labeller = as_labeller(
                   c(`TRUE` = 'After move', `FALSE` = 'Before move'))) +
    scale_x_continuous(breaks = seq(5, 24, 1),
                       labels = c(seq(5, 23, 1), c(0))) +
    scale_y_continuous(labels = scales::percent_format()) +
    theme(panel.grid.minor.x = element_blank(),
          panel.grid.major.x = element_blank(),
          legend.position = 'bottom') +
    labs(title = 'Journeys by hour started',
         x = 'Hour of the day',
         y = 'Number of journeys') +
    guides(fill = FALSE)
```


